Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-09-26 threshold/draft.Rnw
typos

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{ki} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
$[j,k]$ entry in the matrix $V$.)

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\}.
\]
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
restart IRLBA to compute 
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

>>>>

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
$[j,k]$ entry in the matrix $V$.)

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
restart IRLBA to compute 
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

\section{NOTES}

A full/naive thresholded correlation requires $O(n^2 m)$ flops. Let $p$ be the
selected IRLBA dimension. Then the proposed method requires $O(\ell  n  p)$
flops, where $\ell$ is the longest run of ordered entries in $v_1$ that meet
the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$ (not counting
from flops due to IRLBA and the post-processing step of evaluating the
correlation of all the candidate pairs).  One problem with the method as
written is that $\ell$ can a large fraction of n.

Unnecessarily large values of $\ell$ are due to the loose bound
$\sqrt{2(1-t)}/\sigma_1$ used in step 3 of Algorithm\ref{prune} which is only
sharp when $k=1$. General problems may not need to run through differencing
steps up to $\ell$ in step 4 of the algorithm. We can further improve the
efficiency of the algorithm by dynamically reducing $\ell$ in step 4 as
follows.

Equation \ref{vdist} expresses the Euclidean distance between any two columns
of the matrix $A$ in terms of singular components. Write the smallest squared
gap between entries of a particular singular vector $v_h$ as:
\[
\delta_h = \min_{i,j} (e_i^T v_h - e_j^T v_h)^2.
\]
Then note that
\begin{equation}\label{mindist}
\|a_i - a_j\|^2 \ge \sigma_1^2 \delta_1 + \sigma_2^2 \delta_2 + \cdots + \sigma_k^2\delta_k,
\end{equation}
for any indices $i,j$.

Fix $1\le \tilde{i}\le \ell$ and let $\nu \in R^p$ be the vector of minimum values in
each column of the $n\times p$ matrix $D^{\tilde{i}}PV_{1:p}$ from step 4 of
the algorithm.
It follows from Equation \ref{mindist} that
\[
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \ge \nu^Ts_{1:p}
\]
for all $\tilde{j}=1,2,\ldots,n$.

XXX need monotonicity of $\nu^Ts_{1:p}$ XXX
